---
title: "The gofreg package: Perform goodness-of-fit tests for parametric regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gofreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 6
)
```

```{r setup, echo = FALSE, message = FALSE}
options(width = 500, max.print = 500)
set.seed(123)
library(gofreg)
library(ggplot2)
```

## Typical usage

1. Generate or load data
2. Fit a parametric regression model to the given data
3. Compute the p-value of the fitted model

## Example

In this example, we will fit a generalized linear model (GLM) to an artificially
created dataset. It consists of two covariates, one of them normally and the 
other one uniformly distributed, and the response variable following a classical
linear model with normal distribution.

```{r}
set.seed(123)
n  <- 100
x <- cbind(rnorm(n, mean = 3), runif(n, min = 1, max = 10))
model_true <- GLM.new(distr = "normal", linkinv = identity)
params_true <- list(beta = c(2, 6), sd = 1)
y <- model_true$sample_yx(x, params_true)
data <- dplyr::tibble(x = x, y = y)
```

First, we will fit a wrong model to the data. In particular, we exclude the
second covariate.

``` {r}
model_test <- GLM.new(distr = "normal", linkinv = identity)
data_miss <- tibble::tibble(x = data$x[,1], y = data$y)
model_test$fit(data_miss, params_init = list(beta = c(2), sd = 2), inplace = TRUE)
print(model_test$get_params())
```

It can be seen that the variance was estimated to be rather high which is 
reasonable as it includes the part of the variance that could be explained when
taking the second covariate into account. To assess whether the fitted model 
fits to the given data, we perform a bootstrap-based goodness-of-fit test using 
the conditional Kolmogorov test statistic for the marginal distribution of Y.

``` {r}
gt <- GOFTest$new(data = data_miss, model_fitted = model_test, test_stat = CondKolmY$new(), nboot = 100)
print(gt$get_pvalue())
```

As the p-value is very low, the model hypothesis should be rejected. So the test
reveals the mistake in the model assumption. Next, we fit the correct model using
both covariates.

``` {r}
model_test <- GLM.new(distr = "normal", linkinv = identity)
model_test$fit(data, params_init = list(beta = c(1,1), sd = 5), inplace = TRUE)
print(model_test$get_params())
```

The parameters estimates are very close to the true values. The corresponding 
p-value is computed in the following code chunk.

``` {r}
gt <- GOFTest$new(data = data, model_fitted = model_test, test_stat = CondKolmY$new(), nboot = 100)
print(gt$get_pvalue())
```

This time, the p-value is rather high, so the model is cannot be rejected.

## Parametric Regression Models

Here is a list of parametric regression models that are available in the 
`gofreg` package: 

* `NormalGLM`: Generalized linear model with normal distribution
* `GammaGLM`: Generalized linear model with gamma distribution
* `ExpGLM`: Generalized linear model with exponential distribution
* `WeibullGLM`: Generalized linear model with Weibull distribution
* `NegBinomGLM`: Generalized linear model with negative binomial distribution

The package also offers the option to use other user-defined models. For 
instructions on how to implement new models see `vignette("New-Models")`.

## Test Statistics

Here is a list of test statistics that are available in the `gofreg` package: 

* `CondKolmXY`: Conditional Kolmogorov of the joint distribution of $(X,Y)$ defined in Andrews (1997)
* `SICM`: Simulated integrated conditional moment test defined in Bierens & Wang (2012)
* `MEP`: Marked Empirical Process defined in Dikta & Scheer (2021)
* `CondKolmY`: Conditional Kolmogorov of the marginal distribution of $Y$ defined in Kremling's dissertation (2024)
<!-- * `CondKolmbXY`: Conditional Kolmogorov of the joint distribution of $(\beta^T X, Y)$ defined in Kremling & Dikta (2024) -->

The package also offers the option to use other user-defined test statistics. 
For instructions on how to implement new test statistics see
`vignette("New-TestStatistics")`.

## Censored data

The package can also be used to fit parametric regression models and perform 
goodnes-of-fit tests for randmly right-censored survival times $Y$. In this 
case, the `loglik` and `resample` arguments in the `ParamRegrModel$fit()` and
`GOFTest$new()` methods have to be specified. Moreover, the `data` object needs
to be a `data.frame()` with tags "x", "z" and "delta" with $X$ representing the
covariates, $Z = \min(Y, C)$ the censored times and $\delta = 1_{\{Y \le C\}}$ the
censoring indicators. A test statistic for the censored setting is given by
`CondKolmY_RCM`.

Here is an example with artificial data generated from a normal GLM with normally 
distributed censoring times.

``` {r, echo = FALSE, message = FALSE}
set.seed(123)
```

``` {r}
n <- 100
x <- cbind(runif(n), rbinom(n, 1, 0.5))
model <- NormalGLM$new()
y <- model$sample_yx(x, params = list(beta = c(2, 3), sd = 1))
c <- rnorm(n, mean(y) * 1.2, sd(y) * 0.5)
data <- dplyr::tibble(x = x, z = pmin(y, c), delta = as.numeric(y <= c))

model$fit(data, params_init = list(beta = c(1, 1), sd = 3), inplace = TRUE, loglik = loglik_xzd)
print(model$get_params())
```

It can be seen that the estimated parameters are close to the true parameters
$\beta = (2,3)$ and $\sigma = 1$. Now, we compute the corresponding p-value
using the Conditional Kolmogorov test statistic for the marginal distribution
of $Y$ under random censorship (defined in Kremling's dissertation (2024)).

``` {r}
gt <- GOFTest$new(
  data = data, model_fitted = model, test_stat = CondKolmY_RCM$new(), nboot = 100,
  resample = resample_param_cens, loglik = loglik_xzd
)
print(gt$get_pvalue())
```

The p-value is rather high and the model is not rejected which is expected since
we fitted the correct model.
