---
title: "The gofreg package: Perform goodness-of-fit tests for parametric regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gofreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  # collapse = TRUE,
  comment = "",
  fig.width = 6
)
```

```{r setup, echo = FALSE, message = FALSE}
options(width = 500, max.print = 500)
set.seed(123)
library(gofreg)
library(ggplot2)
```

## Typical usage

1. Generate or load data
2. Fit a parametric regression model to the given data
3. Compute the p-value of the fitted model

## Example

In this example, we will fit a Normal and Gaussian Generalized Linear Model 
(GLM) to the `datasets::cars` dataset. Let us first have a look at the data at 
hand.

```{r}
x <- datasets::cars$speed
y <- datasets::cars$dist
data <- dplyr::tibble(x = x, y = y)

pl <- ggplot(data) +
  geom_point(aes(x = x, y = y)) +
  labs(x = "Speed [mph]", y = "Distance [ft]") +
  ggtitle("Stopping distance of cars depending on speed")
print(pl)
```

Now, we fit a GLM with normal distribution to the data. 

``` {r}
model <- GLM.new(distr = "normal", linkinv = identity)
model$fit(data, params_init = list(beta = 3, sd = 2), inplace = TRUE)

params_opt <- model$get_params()
print(params_opt)
```

The resulting regression line (and its 95% confidence band) can be seen in the 
following plot.

``` {r}
pl <- pl + geom_abline(slope = params_opt$beta) +
  geom_abline(intercept = -2 * params_opt$sd, slope = params_opt$beta, linetype = "dashed") +
  geom_abline(intercept = 2 * params_opt$sd, slope = params_opt$beta, linetype = "dashed")
print(pl)
```

To assess whether the fitted model fits to the given data, we perform a 
bootstrap-based goodness-of-fit test using the test statistic given in 
Kremling & Dikta (2024).

``` {r}
gt <- GOFTest$new(data = data, model_fitted = model, test_stat = CondKolmY$new(), nboot = 100)
print(gt$get_pvalue())
```

As the p-value is very low, the model hypothesis should be rejected.
Next, we fit a GLM with gamma distribution instead.

``` {r}
model <- GLM.new(distr = "gamma", linkinv = identity)
model$fit(data, params_init = list(beta = 3, shape = 2), inplace = TRUE)

params_opt <- model$get_params()
print(params_opt)
```

The corresponding p-value is computed in the following code chunk.

``` {r}
gt <- GOFTest$new(data = data, model_fitted = model, test_stat = CondKolmY$new(), nboot = 100)
print(gt$get_pvalue())
```

This time, the p-value is rather high. Hence, a Gamma GLM can be assumed to
describe the given data appropriately. 

## Parametric Regression Models

Here is a list of parametric regression models that are available in the 
`gofreg` package: 

* `NormalGLM`: Generalized linear model with normal distribution
* `GammaGLM`: Generalized linear model with gamma distribution
* `ExpGLM`: Generalized linear model with exponential distribution
* `WeibullGLM`: Generalized linear model with Weibull distribution
* `NegBinomGLM`: Generalized linear model with negative binomial distribution

The package also offers the option to use other user-defined models. For 
instructions on how to implement new models see `vignette("New-Models")`.

## Test Statistics

Here is a list of test statistics that are available in the `gofreg` package: 

* `MEP`: Marked Empirical Process defined in Dikta & Scheer (2021)
* `CondKolmXY`: Conditional Kolmogorov of the joint distribution of $(X,Y)$ defined in Andrews (1997)
* `CondKolmY`: Conditional Kolmogorov of the marginal distribution of $Y$ defined in Kremling & Dikta (2024)
* `CondKolmbXY`: Conditional Kolmogorov of the joint distribution of $(\beta^T X, Y)$ defined in Kremling & Dikta (2024)

The package also offers the option to use other user-defined test statistics. 
For instructions on how to implement new models see
`vignette("New-TestStatistics")`.

## Censored data

The package can also be used to fit parametric regression models and perform 
goodnes-of-fit tests for randmly right-censored survival times $Y$. In this 
case, the `loglik` and `resample` arguments in the `ParamRegrModel$fit()` and
`GOFTest$new()` methods have to be specified. Moreover, the `data` object needs
to be a `data.frame()` with tags "x", "z" and "delta" with $X$ representing the
covariates, $Z = \min(Y, C)$ the censored times and $\delta = 1_{\{Y \le C\}}$ the
censoring indicators.

Here is an example with generated data from a normal GLM with normally 
distributed censoring times.

``` {r, echo = FALSE, message = FALSE}
set.seed(123)
```

``` {r}
n <- 100
x <- cbind(runif(n), rbinom(n, 1, 0.5))
model <- NormalGLM$new()
y <- model$sample_yx(x, params = list(beta = c(2, 3), sd = 1))
c <- rnorm(n, mean(y) * 1.2, sd(y) * 0.5)
data <- dplyr::tibble(x = x, z = pmin(y, c), delta = as.numeric(y <= c))

model$fit(data, params_init = list(beta = c(1, 1), sd = 3), inplace = TRUE, loglik = loglik_xzd)
print(model$get_params())
```

It can be seen that the estimated parameters are close to the true parameters
$\beta = (2,3)$ and $\sigma = 1$. Now, we compute the corresponding p-value
using the Conditional Komlmogorov test statistic for the marginal distribution
of $Y$ under random censorship (defined in Kremling & Dikta (2024)).

``` {r}
gt <- GOFTest$new(
  data = data, model_fitted = model, test_stat = CondKolmY_RCM$new(), nboot = 100,
  resample = resample_param_cens, loglik = loglik_xzd
)
print(gt$get_pvalue())
```

As expected (since we fitted the correct model) the p-value is rather high and 
the model is not rejected. 
